---
title: "OBOBliver — DDEA Course June 2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## How to use this document
1. Work through each section sequentially — later sections depend on earlier ones.
2. For each 💡 exercise, write your code in the provided chunk. The chunk below has the answers.

---

## Setup: Install packages
Run this chunk once before anything else. It takes several minutes on first run.

```{r}
#| label: install
#| eval: false

## if you run this code the first time, install these packages first.



#install.packages(c("tidyverse", "plotly", "sessioninfo"), quiet = TRUE)

#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install(c(
#    "limma",
#    "EnhancedVolcano",
#    "clusterProfiler",
#    "org.Mm.eg.db",
#    "enrichplot"
#), ask = FALSE, update = FALSE)

```

### Load libraries

```{r setup, warning = F, message=F}

library(tidyverse)
library(plotly)
library(limma)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)

```

### Load data
data_norm.tsv is the proteome dataset based on the raw data from https://www.sciencedirect.com/science/article/pii/S1535947622000159
-> liver proteome from WT (lean) and ob/ob mouse (n=4)

this data was already log2 transformed, filtered, imputed, and normalized.

metadata.tsv contains the metadata with column names and grouping.


```{r , warning = F, message=F}
#| label: load-data

data <- read.delim(here::here("data/data_norm.tsv"))
metadata <- read.delim(here::here("data/metadata.tsv"))

```

---

## 1. Check data prerequisites for limma
Limma assumes that data is already log transfomred and normalized, that distributions are comparable across samples, and that there are no (or very few) missing values.

## 💡 Exercise 1: Check the data

Inspect the dataset to verify it meets limma's requirements. 
- Check the data's normality and compare sample distributions
- Check for balanced design and missing values


```{r exercise 1}
#| label: ex1-student

# Your code here
```

```{r exercise 1 answer}
#| label: ex1-answer
#| code-fold: true
#| code-summary: "Show answer"



head(data)
print(metadata)

# Reshape to long format for plotting
data_long <- data |>
    rownames_to_column("protein") |>
    pivot_longer(-protein, names_to = "sample", values_to = "intensity") |>
    left_join(metadata |> rownames_to_column("sample"), by = "sample")

# Boxplot
ggplot(data_long, aes(x = sample, y = intensity)) +
    geom_boxplot(outlier.size = 0.5) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

# Density plot
ggplot(data_long, aes(x = intensity, colour = sample)) +
    geom_density() +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "Density of normalised intensities")

# Missing values
cat("Total NAs:", sum(is.na(data)), "\n")
print(colSums(is.na(data)))

# Samples per group
print(table(metadata$condition))

```

---

## 2. Differential expression analysis with limma

We have two groups: lean mice and ob mice. Start by encoding group membership as a factor:

```{r}
#| label: group-factor

group <- factor(metadata$condition, levels = c("lean", "ob"))

```

## 💡 Exercise 2: Fit the limma model (Approach 1 — no-intercept design)

Use the `~0 +` design so each column of the design matrix represents a group mean directly. Then define an explicit contrast for ob vs lean.

1. Build a design matrix with `model.matrix(~0 + group)` and rename columns
2. Fit the model with `lmFit()`
3. Define the contrast `ob - lean` with `makeContrasts()`
4. Apply the contrast with `contrasts.fit()`, then `eBayes()`
5. Extract results with `topTable()` (all proteins, BH correction)
6. Add a `Gene` column from `rownames()`

Store your final results in `results_approach1`.


```{r exercise 2}
#| label: ex2-student

# your code

```

```{r exercise 2 answer}
#| label: ex2-answer
#| code-fold: true
#| code-summary: "Show answer"

# Step 1: design matrix
design <- model.matrix(~0 + group)
colnames(design) <- c("lean", "ob")

# Step 2: fit model
fit_1 <- lmFit(object = data, design = design)

# Step 3: contrast matrix
contrast_matrix <- makeContrasts(ob - lean, levels = design)

# Step 4: apply contrast + empirical Bayes
fit_2 <- contrasts.fit(fit_1, contrast_matrix)
fit_2 <- eBayes(fit_2)

# Step 5-6: extract results and add Gene column
results_approach1 <- topTable(
    fit_2,
    coef          = "ob - lean",
    number        = Inf,
    adjust.method = "BH"
)
results_approach1$Gene <- rownames(results_approach1)

head(results_approach1)
```

### Reference: Approach 2 — intercept design

There is a simpler alternative for 2-group comparisons that skips `makeContrasts()`:

```{r}
#| label: approach2-reference
#| eval: false

design <- model.matrix(~ group)   # column "groupob" = ob - lean
fit_1  <- lmFit(object = data, design = design)
fit_2  <- eBayes(fit_1)

results_approach2 <- topTable(fit_2,
                               coef          = "groupob",
                               number        = Inf,
                               adjust.method = "BH")

results_approach2$Gene <- rownames(results_approach2)

head(results_approach2)

```

> Both approaches give mathematically identical results. But approach 1 becomes necessary when you have 3+ groups and want to define all pairwise contrasts.

---

## 💡 Exercise 3: P-value distribution

Plot a histogram of the raw p-values from `results_approach1`. What does the shape tell you about whether the model is well-specified?


```{r}
#| label: ex3-answer
#| code-fold: true
#| code-summary: "Show answer"

ggplot(results_approach1, aes(x = P.Value)) +
    geom_histogram(bins = 50, fill = "lightpink", colour = "white") +
    theme_minimal() +
    labs(
        title = "Distribution of raw p-values",
        x     = "Raw p-value",
        y     = "Count"
    )
```




## 3. Visualise limma results

```{r}
#| label: volcano

EnhancedVolcano(results_approach1,
                lab = rownames(results_approach1),
                x   = "logFC",
                y   = "adj.P.Val")




## There are many different options you can add

#  label top hits
top_genes <- rownames(results_approach1[order(results_approach1$adj.P.Val), ])[1:5]

EnhancedVolcano(results_approach1,
                lab = rownames(results_approach1),
                x   = "logFC",
                y   = "adj.P.Val",

              # you can add thresholds
                pCutoff = 0.05,
                FCcutoff = 4,

                selectLab = top_genes,

                # color by adj.p < 0.05
                colCustom = NULL,
                col = c("grey60",      # NS
                        "grey60",      # FC only (same as NS, we don't want this colored)
                        "pink",     # p sig only
                        "purple"),    # p sig + FC

                drawConnectors = TRUE,
                widthConnectors = 0.4,
                colConnectors = "grey40",
                boxedLabels = TRUE,
                labSize = 4,
                pointSize = 1.5)



```

## 💡 Exercise 4: Discussion — interpreting the volcano plot

Discuss with your neighbour: What do  the axes represent? Why do we use -log10 of the adjusted p-value?


---

## 4. Gene Set Enrichment Analysis (GSEA)

GSEA uses the full ranked list of proteins (by logFC), no significance threshold needed.

```{r GSEA, warning = F, message=F}
#| label: gsea

ranked_list <- results_approach1 |>
    arrange(desc(logFC)) |>
    (\(df) setNames(df$logFC, df$Gene))()

gsea <- clusterProfiler::gseGO(
    geneList      = ranked_list,
    ont           = "ALL",          # "CC", "BP", or "MF" for individual ontologies
    keyType       = "SYMBOL",
    minGSSize     = 10,
    maxGSSize     = 500,
    pvalueCutoff  = 0.05,
    verbose       = TRUE,
    OrgDb         = org.Mm.eg.db,
    pAdjustMethod = "BH"
)
```


## 💡 Exercise 5: Visualise GSEA results

Create plots of the GSEA results and discuss what each one shows.

eg you could use:
`dotplot()`, `enrichplot::gseaplot2()`, `enrichplot::emapplot()`, `enrichplot::treeplot()`, `enrichplot::ridgeplot()`
For `emapplot` / `treeplot`: first run `enrichplot::pairwise_termsim()` on the gsea object.

What do the different plots represent? 


```{r exercise 5}
#| label: ex5-student

# Your code here

```

```{r exercise 5 answer, warning = F, message=F}
#| label: ex5-answer
#| code-fold: true
#| code-summary: "Show answer"

# Dotplot split by direction
dotplot(gsea, showCategory = 10, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA GO — all ontologies")

dotplot(gsea, x = "NES", showCategory = 10, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA GO — all ontologies")


# GSEA enrichment curves for top terms
gseaplot2(gsea, geneSetID = 1:3, title = "Top 3 enriched GO terms")

# Similarity-based network and tree (require pairwise_termsim first)
gsea_sim <- enrichplot::pairwise_termsim(gsea)
emapplot(gsea_sim, showCategory = 10)
treeplot(gsea_sim)

# Ridge plot of logFC distributions per term
ridgeplot(gsea, showCategory = 10)
```

---

## 5. Dealing with redundant GO terms

Many GO terms are semantically similar and will appear together. Run a single ontology to make this easier to see:

```{r}
#| label: gsea-mf

gsea_MF <- clusterProfiler::gseGO(
    geneList      = ranked_list,
    ont           = "MF",
    keyType       = "SYMBOL",
    minGSSize     = 10,
    maxGSSize     = 500,
    pvalueCutoff  = 0.05,
    verbose       = TRUE,
    OrgDb         = org.Mm.eg.db,
    pAdjustMethod = "BH"
)

dotplot(gsea_MF)
```

## 💡 Exercise 6: Simplify redundant terms

Do you see redundant GO terms in the dotplot above?

Use `clusterProfiler::simplify()` to reduce redundancy. Try adjusting the `cutoff` parameter (default 0.7 — lower values are more aggressive). Compare the dotplots before and after.


```{r exercise 6}
#| label: ex6-student

# Your code here
```

```{r exercise 6 answer}
#| label: ex6-answer
#| code-fold: true
#| code-summary: "Show answer"

gsea_MF_simplified <- clusterProfiler::simplify(gsea_MF, cutoff = 0.7)

dotplot(gsea_MF_simplified) +
    ggtitle("After simplify() — cutoff 0.7")

# How many terms remain?
n_before <- sum(gsea_MF@result$p.adjust < 0.05)
n_after  <- sum(gsea_MF_simplified@result$p.adjust < 0.05)
cat(sprintf("Terms before: %d  |  after: %d\n", n_before, n_after))

# Try a more aggressive cutoff
gsea_MF_strict <- clusterProfiler::simplify(gsea_MF, cutoff = 0.4)
dotplot(gsea_MF_strict) +
    ggtitle("After simplify() — cutoff 0.4 (more aggressive)")
```

---

## 6. KEGG pathway enrichment

KEGG requires Entrez IDs rather than gene symbols.

```{r}
#| label: gsea-kegg

gene_entrez <- clusterProfiler::bitr(
    results_approach1$Gene,
    fromType = "SYMBOL",
    toType   = "ENTREZID",
    OrgDb    = org.Mm.eg.db
)

ranked_kegg <- results_approach1 |>
    left_join(gene_entrez, by = c("Gene" = "SYMBOL")) |>
    filter(!is.na(ENTREZID)) |>
    arrange(desc(logFC)) |>
    (\(df) setNames(df$logFC, df$ENTREZID))()

gsea_kegg <- clusterProfiler::gseKEGG(
    geneList      = ranked_kegg,
    organism      = "mmu",
    minGSSize     = 10,
    maxGSSize     = 500,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "BH",
    verbose       = TRUE
)

dotplot(gsea_kegg, showCategory = 10, split = ".sign", font.size = 8) +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA-KEGG") +
    theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))
```

---

## 7. Over-Representation Analysis (ORA)

ORA tests whether a predefined gene set is over-represented among your significant proteins compared to a background

```{r}
#| label: ora

sig        <- results_approach1 |> filter(adj.P.Val < 0.05) |> pull(Gene)
background <- results_approach1 |> pull(Gene)

ora <- clusterProfiler::enrichGO(
    gene          = sig,
    universe      = background,
    OrgDb         = org.Mm.eg.db,
    keyType       = "SYMBOL",
    ont           = "ALL",
    minGSSize     = 10,
    maxGSSize     = 500,
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05
)
```


## 💡 Exercise 7: Visualise ORA and compare to GSEA

1. Create a dotplot of the ORA results
2. Run ORA separately for upregulated and downregulated proteins
3. Discuss with your neighbour:
   - What is the key conceptual difference between ORA and GSEA?
   - Do the two methods agree on the enriched pathways?
   - What are the pros and cons of each?


```{r exercise 7}
#| label: ex7-student

# Your code here

# Hint: filter results_approach1 for adj.P.Val < 0.05 and logFC > 0 (up) or < 0 (down),
# then run enrichGO separately for each direction

```


```{r exercise 7 answer}
#| label: ex7-answer
#| code-fold: true
#| code-summary: "Show answer"

# Dotplot of combined ORA
dotplot(ora, showCategory = 10) +
    ggtitle("ORA GO — all significant proteins")

# Upregulated proteins only
up_genes <- results_approach1 |>
    filter(adj.P.Val < 0.05, logFC > 0) |>
    pull(Gene)

ora_up <- clusterProfiler::enrichGO(
    gene          = up_genes,
    universe      = background,
    OrgDb         = org.Mm.eg.db,
    keyType       = "SYMBOL",
    ont           = "ALL",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05
)
dotplot(ora_up, showCategory = 10) + ggtitle("ORA — Upregulated in ob")

# Downregulated proteins only
down_genes <- results_approach1 |>
    filter(adj.P.Val < 0.05, logFC < 0) |>
    pull(Gene)

ora_down <- clusterProfiler::enrichGO(
    gene          = down_genes,
    universe      = background,
    OrgDb         = org.Mm.eg.db,
    keyType       = "SYMBOL",
    ont           = "ALL",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05
)
dotplot(ora_down, showCategory = 10) + ggtitle("ORA — Downregulated in ob")
```


## 8. KEGG ORA

```{r}
#| label: ora-kegg

sig_entrez        <- gene_entrez |> filter(SYMBOL %in% sig) |> pull(ENTREZID)
background_entrez <- gene_entrez$ENTREZID

ora_kegg <- clusterProfiler::enrichKEGG(
    gene          = sig_entrez,
    universe      = background_entrez,
    organism      = "mmu",
    minGSSize     = 10,
    maxGSSize     = 500,
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05
)

dotplot(ora_kegg, showCategory = 10, font.size = 8) +
    ggtitle("ORA-KEGG") +
    theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))
```





## EXTRA EXERCISES 
if you have the time, go through the limma user manual and look into how to write the code for more complex designs for limma analysis.


## 💡 Exercise 8: Discussion — 3-group design

Imagine you had three groups: lean, ob, and diabetic. How would you perform the limma DEA? discuss with your neighbour.


## 💡 Exercise 9: Discussion — paired design

imagine you had two groups (treated and control) and one measurement of each subject pre- and post- intervetion - hence a paired design. How can you now run limma in a paired design?



