---
title: "Proteomics Data Visualization"
subtitle: "DDEA Proteomics Course — Hands-On Tutorial"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    code-link: true
    df-print: paged
    fig-width: 7
    fig-height: 5
execute:
  warning: false
  message: false
  echo: true
editor: source
---

```{r setup}
#| include: false

# Dimensionality reduction & embedding
library(Seurat)             # UMAP + FeaturePlot workflow

# Pathway enrichment
library(clusterProfiler)    #GSEA
library(org.Hs.eg.db)       #Human gene annotations

# Differential analysis & visualization
library(limma)
library(plotly)             # interactive volcano via ggplotly
library(pheatmap)           #Heatmap and clustering

# Trajectory / neighbourhood testing
library(miloR)

# General data wrangling:
library(tidyverse)
```

------------------------------------------------------------------------

## Learning Objectives

By the end of this session you will be able to:

1.  Describe the structure of a single-fiber proteomics dataset and access its intensity matrix and sample metadata.
2.  Run and interpret linear (PCA) and non-linear (UMAP) dimensionality-reduction methods on protein-intensity data.
3.  Extract latent vectors from a low-dimensional embedding and relate them to biological covariates.
4.  Use neighbourhood-based testing (MILO) to detect differential abundance across phenotypic groups in single-fiber space.
5.  Visualize the results of a differential abundance analysis using boxplots, volcano plots, UMAP feature plots, and covariate–intensity associations.
6.  Build heatmaps of differentially abundant proteins, cluster them into co-regulated modules, and link those modules to biological pathways.

------------------------------------------------------------------------

# Dataset: Single-Fiber Human Muscle Proteomics

::: callout-note
This section is a placeholder — update with the actual cohort description, acquisition method, and any preprocessing already applied before the session.
:::

The example dataset contains protein intensities for **thousands of single muscle fibers** dissected from human donors. The data is distributed as an `.rds` file containing a **named list** with (at least) two objects:

| Object | Type | Rows | Columns | Notes |
|---------------|---------------|---------------|---------------|---------------|
| `intensities` | `matrix` / `data.frame` | proteins | single fibers | log2-transformed, normalized intensities |
| `metadata` | `data.frame` | single fibers | sample-level annotations | donor ID, fiber type, condition, etc. |

> **TODO**: replace with the exact object names, dimensions, normalization scheme, and donor / condition design once finalized.

```{r load-data}
#| eval: true

# Load the prepared dataset
fiber_data <- readRDS("data/single_fiber_proteomics.rds")

# Inspect structure
str(fiber_data, max.level = 1)

# Convenience handles
intensities <- fiber_data$intensities   # proteins x fibers
metadata    <- fiber_data$metadata      # one row per fiber

dim(intensities)
head(metadata)
```

> **Exercise 1.1**: Inspect `metadata`. Which variables are biological (donor, fiber type, …) and which are technical (plate, run order, …)? Why does this distinction matter for the analyses that follow?

<details>

<summary><strong>Exercise 1.1 — possible answer</strong></summary>

TODO — fill in once the metadata fields are finalized. The point to make: biological covariates are what we *want* the embedding to reflect; technical covariates are confounders we want to rule out as drivers of the structure we see.

</details>

------------------------------------------------------------------------

# Stage 1: Dimensionality Reduction & Clustering

## Why reduce dimensionality?

A single-fiber proteomics matrix has thousands of proteins measured across thousands of fibers — too many dimensions to visualize directly, and most pairs of proteins are correlated. Dimensionality reduction projects the data into a small number of axes that capture the strongest sources of variation, letting us see global structure (fiber types, donors, conditions) in two or three dimensions.

| Method | Type | What it preserves | Use it for |
|-----------------|-----------------|---------------------|------------------|
| PCA | Linear | Global variance; directions of maximum spread | Detecting dominant gradients / batches |
| UMAP | Non-linear | Local neighbourhood structure | Revealing discrete populations / clusters |

## 1.1 Principal Component Analysis (PCA)

PCA finds orthogonal linear combinations of proteins that maximize variance. PC1 captures the strongest gradient in the data, PC2 the next strongest orthogonal to PC1, and so on. Each fiber gets a score on each PC; each protein gets a loading.

```{r pca}
#| eval: true

# prcomp expects observations in rows and features in columns,
# so we transpose: rows = fibers, columns = proteins.
# It does NOT tolerate NA, so we mean-impute first (row-wise on the original matrix).
imp <- intensities |> 
    PhosR::tImpute() |> 
    as.data.frame()

pca_res <- prcomp(t(imp), scale. = TRUE)

# --- Scree plot: variance explained per PC ---
var_exp <- (pca_res$sdev^2) / sum(pca_res$sdev^2)

tibble(PC = seq_along(var_exp), var_exp = var_exp) %>%
  slice_head(n = 20) %>%
  ggplot(aes(PC, var_exp)) +
  geom_col() +
  scale_x_continuous(breaks = 1:20) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(y = "Variance explained")

# --- Attach PC1 / PC2 back to metadata so downstream chunks can use them ---
metadata <- metadata %>%
  left_join(
    as_tibble(pca_res$x, rownames = "fiber_id") %>% select(fiber_id, PC1, PC2),
    by = "fiber_id"
  )

# --- PCA scatter: PC1 vs PC2, coloured by time, shaped by subject ---
ggplot(metadata, aes(PC1, PC2, colour = time)) +
  geom_point(size = 0.8, alpha = 0.7) +
  theme_minimal()
```

> **Exercise 1.2**: Look at the scree plot. How many PCs together explain the major part of the variance?

<details>

<summary><strong>Exercise 1.2 — possible answer</strong></summary>

PC1 alone explains a great % of variance, almost three times more than PC2. This data set has an important amount of information captured by PC1.

</details>

> **Exercise 1.3**: Time doesn't seem to be explained by either PC1 or PC2.
>
> 1.  What about subject? Can either PC1 or PC2 explain subject variability? Color the PCA by subject and find out!
>
> 2.  Then, try extracting PC3 and plotting it against PC1. Does it capture the effect of time better? Look at the previous code chunk for inspiration!

<details>

<summary><strong>Exercise 1.3.1 — solution</strong></summary>

```{r ex-1-3-1}
#| eval: true

ggplot(metadata, aes(PC1, PC2, colour = subject_id)) +
  geom_point(size = 0.8, alpha = 0.7) +
  theme_minimal()
```

</details>

<details>

<summary><strong>Exercise 1.3.2 — solution</strong></summary>

```{r ex-1-3-2}
#| eval: true

metadata_exercise <- metadata %>%
  left_join(
    as_tibble(pca_res$x, rownames = "fiber_id") %>% select(fiber_id, PC3),
    by = "fiber_id"
  )

# --- PCA scatter: PC1 vs PC3, coloured by time ---
ggplot(metadata_exercise, aes(PC1, PC3, colour = time)) +
  geom_point(size = 0.8, alpha = 0.7) +
  theme_minimal()
```

</details>

## 1.2 Extracting vectors of information from PCA

The fiber-level PC scores live in `pca_res$x` (one row per fiber, one column per PC). Protein loadings live in `pca_res$rotation`. These two matrices are what you carry forward:

-   We already joined the PC score/sample for PC1 and PC2 back to `metadata` for downstream modelling.
-   These can be tested for associations with protein intensities using linear models, GAMs, etc.(`lm( ~ intensity + PCx)`).
-   By extracting the loadings for each PC, we can see which proteins are the greatest drivers of that PC.

```{r pca-vectors}
#| eval: true

pc_scores <- as_tibble(pca_res$x, rownames = "fiber_id") %>%
  left_join(metadata |> 
                dplyr::select(!c(PC1, PC2)), by = "fiber_id")

# Quick association screen: which covariates drive which PCs?
# Example: PC1 vs. time
summary(lm(PC1 ~ time, data = pc_scores))
summary(lm(PC3 ~ time, data = pc_scores))

# PC1 is associated with time, however, PC3 shows a greater association
```


PCA loadings can be extracted from `pca_res$rotation` and used, for example, as an input for a ranked GSEA test. The ranking statistic here is the **loading of each protein on PC1** — a signed weight, where strongly positive and strongly negative loadings both indicate proteins that drive the axis (just in opposite directions). `clusterProfiler::gseGO` walks down the ranked list and tests whether each GO Biological Process term is enriched at one extreme.

```{r pca-loadings}
#| eval: true

# --- 1. Extract PC1 loadings as a named, ranked vector ---
PC1_loadings <- sort(pca_res$rotation[, "PC1"], decreasing = TRUE)
head(PC1_loadings, 10)   # most positive
tail(PC1_loadings, 10)   # most negative

# --- 2. Ranked GSEA against GO Biological Process ---
# Gene symbols are the keys, so we set keyType = "SYMBOL".
# On Windows, the default BiocParallel backend (SnowParam) sometimes fails
# to serialize the permutation step. Force serial execution to avoid this.
BiocParallel::register(BiocParallel::SerialParam())

set.seed(42)
gse_PC1 <- clusterProfiler::gseGO(
  geneList     = PC1_loadings,
  OrgDb        = org.Hs.eg.db::org.Hs.eg.db,
  keyType      = "SYMBOL",
  ont          = "BP",
  minGSSize    = 15,
  maxGSSize    = 500,
  pvalueCutoff = 0.05,
  verbose      = FALSE
)

# --- 3. Tidy results and pick top terms (10 up, 10 down by padj) ---
gse_PC1_df <- as_tibble(gse_PC1@result) %>% arrange(p.adjust)

top_terms <- bind_rows(
  gse_PC1_df %>% filter(NES > 0) %>% slice_head(n = 10),
  gse_PC1_df %>% filter(NES < 0) %>% slice_head(n = 10)
)

# --- 4. Dot plot ---
top_terms %>%
  ggplot(aes(x = NES,
             y = forcats::fct_reorder(Description, NES),
             size   = -log10(p.adjust),
             colour = NES)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_gradient2(low = "steelblue", mid = "grey80", high = "firebrick",
                         midpoint = 0) +
  theme_minimal() +
  labs(title = "GO:BP terms ranked by PC1 loading",
       x = "Normalized enrichment score (NES)",
       y = NULL,
       size = "-log10(p.adjust)")
```

> **Exercise 1.4**: PC1 captures the dominant axis of variation in the data, but the other PCs may carry biology too. Repeat the ranked GSEA for **PC2** and **PC3** and inspect their dot plots. What is the dominant biology associated with each? Do any pathways appear on multiple PCs (and what would that mean)?

<details>

<summary><strong>Exercise 1.4 — solution</strong></summary>

The workflow is identical to PC1 — only the column we pull from `pca_res$rotation` changes. Wrapping it in a small helper keeps the code clean and produces one dot plot per PC:

```{r ex-1-4}
#| eval: true

plot_pc_gsea_go <- function(pc, seed = 42) {

  loadings <- sort(pca_res$rotation[, pc], decreasing = TRUE)

  BiocParallel::register(BiocParallel::SerialParam())
  set.seed(seed)
  gse <- clusterProfiler::gseGO(
    geneList     = loadings,
    OrgDb        = org.Hs.eg.db::org.Hs.eg.db,
    keyType      = "SYMBOL",
    ont          = "BP",
    minGSSize    = 15,
    maxGSSize    = 500,
    pvalueCutoff = 0.05,
    verbose      = FALSE
  )

  top <- as_tibble(gse@result) %>%
    arrange(p.adjust) %>%
    {dplyr::bind_rows(
       dplyr::filter(., NES > 0) %>% slice_head(n = 10),
       dplyr::filter(., NES < 0) %>% slice_head(n = 10)
    )}

  ggplot(top, aes(x = NES,
                  y = forcats::fct_reorder(Description, NES),
                  size   = -log10(p.adjust),
                  colour = NES)) +
    geom_point() +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
    scale_colour_gradient2(low = "steelblue", mid = "grey80",
                           high = "firebrick", midpoint = 0) +
    theme_minimal() +
    labs(title = paste("GO:BP terms ranked by", pc, "loading"),
         x = "Normalized enrichment score (NES)",
         y = NULL,
         size = "-log10(p.adjust)")
}

plot_pc_gsea_go("PC2")
plot_pc_gsea_go("PC3")
```

</details>

## 1.3 Uniform Manifold Approximation and Projection (UMAP)

UMAP is a non-linear embedding that prioritizes preserving **local** neighbourhood structure. It is well-suited for revealing discrete populations (e.g. fast vs. slow fibers, hybrid fibers) that may not separate along a single linear axis.

We use the **Seurat workflow**, which packages PCA → neighbour graph → UMAP into a single object and provides `DimPlot` / `FeaturePlot` helpers we will reuse later.

```{r umap}
#| eval: true

# Seurat does not tolerate NA — feed it the mean-imputed matrix from the PCA step.
# Rownames of the metadata data.frame must match the column names of the matrix.
md <- metadata %>% column_to_rownames("fiber_id")

seu <- Seurat::CreateSeuratObject(counts = imp, meta.data = md)

# Intensities are already log-transformed and normalized, so we skip
# NormalizeData. Treat every protein as a "variable feature".
Seurat::VariableFeatures(seu) <- rownames(seu)
seu[["RNA"]]$data <- seu[["RNA"]]$counts

# --- Reuse the prcomp PCs from Section 1.1 as Seurat's PCA reduction ---
# This keeps the whole course pipeline (PCA -> UMAP -> MILO) on a single set
# of coordinates, instead of recomputing PCA per package.
pc_emb <- pca_res$x[, 1:30]
pc_ldg <- pca_res$rotation[, 1:30]
colnames(pc_emb) <- paste0("PC_", seq_len(ncol(pc_emb)))
colnames(pc_ldg) <- paste0("PC_", seq_len(ncol(pc_ldg)))

seu[["pca"]] <- Seurat::CreateDimReducObject(
  embeddings = pc_emb[colnames(seu), ],
  loadings   = pc_ldg,
  assay      = Seurat::DefaultAssay(seu),
  key        = "PC_"
)

set.seed(42)
seu <- Seurat::RunUMAP(seu, dims = 1:20, verbose = FALSE)

Seurat::DimPlot(seu, reduction = "umap", group.by = "time", pt.size = 0.4)
```

The embedding itself is stored in `Embeddings(seu, "umap")` if you need a `data.frame` for custom plots:

```{r umap-embeddings}
#| eval: true

umap_df <- Seurat::Embeddings(seu, "umap") %>%
  as_tibble(rownames = "fiber_id") %>%
  dplyr::rename(UMAP1 = umap_1, UMAP2 = umap_2) %>%
  left_join(metadata, by = "fiber_id")
```

> **Exercise 1.5**: Compare the UMAP to the PC1 vs PC2 and PC1 vs PC3 scatter plots you made in Exercises 1.3.1 and 1.3.2. Does any structure become visible in the UMAP that the linear projections missed? Can you tell from the UMAP alone whether fibers separate by `time`?

<details>

<summary><strong>Exercise 1.5 — possible answer</strong></summary>

UMAP often resolves discrete sub-populations into well-separated islands that the linear PCs only show as smooth gradients along a single axis. In this dataset the fiber-type structure (slow / fast) tends to be clearer in UMAP. The `time` effect is more subtle and is usually not what drives the UMAP layout — exactly as in the PCA, the dominant axis is fiber type rather than the perturbation.

</details>

> **Exercise 1.6**: UMAP has two knobs that change the layout substantially. Try varying each in turn and re-plot.
>
> 1.  Re-run UMAP with `n.neighbors = 5` and `n.neighbors = 100`. How does each value emphasize local vs. global structure?
>
> 2.  Re-run UMAP using only the first 5 PCs (`dims = 1:5`) and 30 PCs (`dims = 1:30`). What does the input dimensionality control?

<details>

<summary><strong>Exercise 1.6.1 — solution</strong></summary>

A Seurat object can hold multiple UMAP reductions side-by-side via `reduction.name`, so we don't need to rebuild the object:

```{r ex-1-6-1}
#| eval: true

seu <- Seurat::RunUMAP(seu, dims = 1:20, n.neighbors = 5,
                       reduction.name = "umap_n5",   verbose = FALSE)
seu <- Seurat::RunUMAP(seu, dims = 1:20, n.neighbors = 100,
                       reduction.name = "umap_n100", verbose = FALSE)

Seurat::DimPlot(seu, reduction = "umap_n5",   group.by = "time", pt.size = 0.4) +
  ggplot2::ggtitle("n.neighbors = 5")

Seurat::DimPlot(seu, reduction = "umap_n100", group.by = "time", pt.size = 0.4) +
  ggplot2::ggtitle("n.neighbors = 100")
```

Small `n.neighbors` emphasizes very local structure (more, smaller islands; noisier layout). Large `n.neighbors` emphasizes global structure (fewer, broader clusters; closer to a PCA-like view).

</details>

<details>

<summary><strong>Exercise 1.6.2 — solution</strong></summary>

```{r ex-1-6-2}
#| eval: true

seu <- Seurat::RunUMAP(seu, dims = 1:5,
                       reduction.name = "umap_d5",  verbose = FALSE)
seu <- Seurat::RunUMAP(seu, dims = 1:30,
                       reduction.name = "umap_d30", verbose = FALSE)

Seurat::DimPlot(seu, reduction = "umap_d5",  group.by = "time", pt.size = 0.4) +
  ggplot2::ggtitle("dims = 1:5")

Seurat::DimPlot(seu, reduction = "umap_d30", group.by = "time", pt.size = 0.4) +
  ggplot2::ggtitle("dims = 1:30")
```

`dims` controls how many PCs feed into the neighbour graph. With only 5 PCs, UMAP sees a low-resolution summary of the data — fine for the dominant axis, but it discards the variation captured by later PCs. With 30, more biology and more noise both enter; the layout is richer but can become unstable, especially if many of those PCs carry mostly noise. The scree plot from Exercise 1.2 is a good guide — pick `dims` up to where the variance curve flattens.

</details>

## 1.4 Clustering and fiber-type annotation

The UMAP shows several distinct populations. We can now (1) hard-cluster the fibers on the kNN graph we already built, (2) label each cluster as **fast** or **slow** based on the dominant myosin heavy chain, and (3) carry that label downstream into the MILO model. Canonical markers in human skeletal muscle:

| Marker | Fiber type            |
|--------|-----------------------|
| MYH7   | Type I (slow oxidative)        |
| MYH2   | Type IIa (fast oxidative/glycolytic) |

```{r clustering}
#| eval: true

# FindNeighbors uses the PCA reduction we already injected; FindClusters runs
# Louvain on that graph.
seu <- Seurat::FindNeighbors(seu, dims = 1:20, verbose = FALSE)
seu <- Seurat::FindClusters(seu, resolution = 0.5, verbose = FALSE)

Seurat::DimPlot(seu, reduction = "umap", group.by = "seurat_clusters",
                pt.size = 0.4, label = TRUE) +
  ggplot2::ggtitle("UMAP clusters")
```

```{r myosin-featureplot}
#| eval: true

# Confirm which clusters are slow (MYH7-high) vs fast (MYH2-high)
Seurat::FeaturePlot(seu, features = c("MYH7", "MYH2"),
                    reduction = "umap", pt.size = 0.4, order = TRUE) &
  ggplot2::scale_colour_viridis_c()
```

```{r assign-fiber-type}
#| eval: true

# Per-cluster mean intensity for the two myosin markers
cluster_myh <- tibble::tibble(
    fiber_id = colnames(seu),
    cluster  = as.character(seu$seurat_clusters),
    MYH7     = intensities["MYH7", colnames(seu)],
    MYH2     = intensities["MYH2", colnames(seu)]
  ) %>%
  group_by(cluster) %>%
  summarise(
    MYH7 = mean(MYH7, na.rm = TRUE),
    MYH2 = mean(MYH2, na.rm = TRUE),
    n    = dplyr::n(),
    .groups = "drop"
  ) %>%
  mutate(fiber_type = ifelse(MYH2 > MYH7, "fast", "slow"))

cluster_myh

# Push the fiber_type label back onto seu and metadata
fiber_type_map <- setNames(cluster_myh$fiber_type, cluster_myh$cluster)
seu$fiber_type <- unname(fiber_type_map[as.character(seu$seurat_clusters)])

metadata <- metadata %>%
  left_join(
    tibble::tibble(fiber_id   = colnames(seu),
                   fiber_type = seu$fiber_type),
    by = "fiber_id"
  )

# Sanity check on the UMAP
Seurat::DimPlot(seu, reduction = "umap", group.by = "fiber_type",
                pt.size = 0.4) +
  ggplot2::ggtitle("UMAP coloured by fiber type")
```

> **Exercise 1.7**: ACTN3 is another well-established fast-fiber marker. Use `FeaturePlot` to confirm that the clusters you labelled as `"fast"` also show elevated ACTN3 — if they don't, your fiber-type assignment may need rechecking.

<details>

<summary><strong>Exercise 1.7 — solution</strong></summary>

```{r ex-1-7}
#| eval: true

Seurat::FeaturePlot(seu, features = "ACTN3", reduction = "umap",
                    pt.size = 0.4, order = TRUE) +
  ggplot2::scale_colour_viridis_c()
```

</details>

## 1.5 Differential abundance in embedding space — MILO

MILO tests for differential abundance across neighbourhoods of a kNN graph built on the embedding, instead of treating every fiber as independent. It avoids hard clustering and gives a continuous map of where two groups differ in fiber-state space.

Now that every fiber carries a `fiber_type` label we can refine the design: define the biological sample as `subject × time × fiber_type` (4 samples per subject, 32 total) and ask **two** questions of the data:

1. **Main effect** — block on subject and fiber type and test the time effect averaged across both fiber types.
2. **Stratified by fiber type** — fit a separate `post − pre` contrast within fast fibers and within slow fibers, then **count and compare** how many neighbourhoods are significant in each. If one fiber type responds more to the perturbation, it should show more (or larger-magnitude) significant neighbourhoods.

```{r milo}
#| eval: true

# --- Build a local metadata table with the biological "sample" column ---
# MILO counts how many fibers from each sample fall in each neighbourhood,
# so `sample` must group fibers, NOT identify them. With fiber_type now
# available, each (subject, time, fiber_type) cell is one biological sample.
#
# We pull `fiber_type` directly from `seu` so this chunk works regardless of
# whether `metadata` was reset by re-running an upstream chunk.
stopifnot("fiber_type" %in% colnames(seu@meta.data))

md <- metadata %>%
  mutate(sample_id = paste(subject_id, time, fiber_type, sep = "_")) %>%
  as.data.frame()
rownames(md) <- md$fiber_id

# Column order of the SCE assay must match metadata row order so colData lines up
imp_mat <- as.matrix(imp)[, md$fiber_id]

sce <- SingleCellExperiment::SingleCellExperiment(
  assays  = list(logcounts = imp_mat),
  colData = md
)

# --- Reuse the PCs and UMAP coordinates from Sections 1.1 and 1.3 ---
# Sharing coordinates across all three steps (PCA -> UMAP -> MILO) means the
# neighbourhood map shown by plotNhoodGraphDA() is the SAME embedding the
# students saw earlier in the Seurat DimPlot, not a re-computation that
# would shuffle / rotate the layout.
SingleCellExperiment::reducedDim(sce, "PCA")  <- pca_res$x[colnames(sce), 1:30]
SingleCellExperiment::reducedDim(sce, "UMAP") <-
  Seurat::Embeddings(seu, "umap")[colnames(sce), ]

milo_obj <- miloR::Milo(sce)
milo_obj <- miloR::buildGraph(milo_obj,  k = 30, d = 20, reduced.dim = "PCA")
milo_obj <- miloR::makeNhoods(milo_obj,  prop = 0.1, k = 30, d = 20,
                              refined = TRUE, reduced_dims = "PCA")
milo_obj <- miloR::countCells(milo_obj,  meta.data = md, sample = "sample_id")
milo_obj <- miloR::buildNhoodGraph(milo_obj)

# --- Design table: one row per sample_id ---
design <- md %>%
  distinct(sample_id, subject_id, time, fiber_type) %>%
  as.data.frame()
rownames(design) <- design$sample_id

# Set explicit reference levels so the contrast names are predictable
design$time       <- factor(design$time,       levels = c("pre",  "post"))
design$fiber_type <- factor(design$fiber_type, levels = c("fast", "slow"))

# Combined cell-means factor (fast_pre / fast_post / slow_pre / slow_post).
# This avoids colon-named interaction terms, which makeContrasts() can't parse.
design$group <- factor(
  paste(design$fiber_type, design$time, sep = "_"),
  levels = c("fast_pre", "fast_post", "slow_pre", "slow_post")
)

# 1) MAIN effect: post vs pre, blocking on subject and fiber_type
da_time <- miloR::testNhoods(milo_obj,
                             design    = ~ subject_id + fiber_type + time,
                             design.df = design)

# 2) Time effect WITHIN each fiber type, as two separate contrasts on the
#    `group` factor. We drop the intercept (`~ 0 + group + ...`) so that all
#    four group levels show up as named coefficients — otherwise the reference
#    level (`fast_pre`) is absorbed into the intercept and the contrast
#    `groupfast_post - groupfast_pre` would fail. Same neighbourhoods for both
#    contrasts, so the resulting counts are directly comparable.
da_fast <- miloR::testNhoods(
  milo_obj,
  design          = ~ 0 + group + subject_id,
  design.df       = design,
  model.contrasts = "groupfast_post - groupfast_pre"
)

da_slow <- miloR::testNhoods(
  milo_obj,
  design          = ~ 0 + group + subject_id,
  design.df       = design,
  model.contrasts = "groupslow_post - groupslow_pre"
)

# --- Count significant neighbourhoods per fiber type ---
da_counts <- tibble::tibble(
  fiber_type = c("fast", "slow"),
  n_total    = c(nrow(da_fast), nrow(da_slow)),
  n_sig      = c(sum(da_fast$SpatialFDR < 0.1, na.rm = TRUE),
                 sum(da_slow$SpatialFDR < 0.1, na.rm = TRUE)),
  n_up       = c(sum(da_fast$SpatialFDR < 0.1 & da_fast$logFC > 0, na.rm = TRUE),
                 sum(da_slow$SpatialFDR < 0.1 & da_slow$logFC > 0, na.rm = TRUE)),
  n_down     = c(sum(da_fast$SpatialFDR < 0.1 & da_fast$logFC < 0, na.rm = TRUE),
                 sum(da_slow$SpatialFDR < 0.1 & da_slow$logFC < 0, na.rm = TRUE))
)

da_counts

# --- Bar chart: significant neighbourhoods per fiber type ---
da_counts %>%
  tidyr::pivot_longer(c(n_up, n_down),
                      names_to = "direction", values_to = "n") %>%
  mutate(direction = recode(direction,
                            n_up   = "post > pre",
                            n_down = "pre > post")) %>%
  ggplot(aes(fiber_type, n, fill = direction)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(x = "Fiber type",
       y = "Significant neighbourhoods (SpatialFDR < 0.1)",
       fill = NULL)

# --- The MILO maps themselves: where in fiber-state space each type changes ---
miloR::plotNhoodGraphDA(milo_obj, da_time, layout = "UMAP", alpha = 0.1) +
  ggplot2::ggtitle("Main time effect (post vs. pre)")

miloR::plotNhoodGraphDA(milo_obj, da_fast, layout = "UMAP", alpha = 0.1) +
  ggplot2::ggtitle("post vs. pre — fast fibers")

miloR::plotNhoodGraphDA(milo_obj, da_slow, layout = "UMAP", alpha = 0.1) +
  ggplot2::ggtitle("post vs. pre — slow fibers")
```

> **Exercise 1.8**: Look at the `da_counts` table and the bar chart.
>
> 1.  Which fiber type has more significant neighbourhoods between `pre` and `post`?
>
> 2.  In each fiber type, are the changes mostly in one direction (more fibers in `post`) or balanced between gain and loss?
>
> 3.  Cross-check against the two MILO maps — do the significant neighbourhoods for the fast contrast sit in the fast-dominant region of the UMAP, and likewise for slow? If a contrast lights up neighbourhoods that are dominated by the *other* fiber type, what does that tell you about how confident the count comparison should make you?

<details>

<summary><strong>Exercise 1.8 — possible answer</strong></summary>

TODO — depends on the dataset.

The interpretation pattern to look for:

-   If one fiber type has many more significant neighbourhoods than the other, that fiber type is more responsive to the perturbation. Pay attention to the *direction* too — `post > pre` and `pre > post` counts can tell you whether the population is expanding or shrinking in fiber-state space.
-   A contrast (e.g. fast `post − pre`) tests counts of *fast* fibers in every neighbourhood. Neighbourhoods that are essentially all-slow have near-zero fast counts and won't pass the significance threshold no matter what — so the count comparison is fair only where each fiber type has enough cells to power the test. The two MILO maps make this visible: if the `fast` map only shows significant nhoods in the fast-dominant region, the test is doing the right thing.

</details>

------------------------------------------------------------------------

# Stage 2: Visualizing Differential Abundance Results

The model fitted upstream (a per-protein GAM with `s(PC1, by = time)` and a subject random effect) returns, for every gene, three p-values that summarize different biological questions:

| Column | Question it answers |
|-------------------|-----------------------------------------------------|
| `Estimate_time1` | Effect size (log2FC) of `post` vs. `pre`, averaged over PC1 |
| `Pr(>|t|)_time1` | Is the protein differentially abundant between `pre` and `post`? |
| `p_val_fiber_type` | Is the protein's intensity associated with PC1 (i.e. fiber type)? |
| `p_val_interaction` | Does the time effect depend on PC1 (i.e. fiber-type-specific response)? |

Results are stored as a **named list** — one entry per gene, each entry containing a one-row tibble `model_results`. We start by flattening it into a tidy data frame for plotting.

```{r da-input}
#| eval: true

model_list <- readRDS("results/model_results.rds")

# Flatten the named list into one row per protein
da_results <- dplyr::bind_rows(lapply(model_list, `[[`, "model_results")) %>%
  dplyr::mutate(
    adj_p_time        = p.adjust(`Pr(>|t|)_time1`,  method = "BH"),
    adj_p_fiber_type  = p.adjust(p_val_fiber_type,  method = "BH"),
    adj_p_interaction = p.adjust(p_val_interaction, method = "BH")
  )

head(da_results)
```

A quick orientation before we start visualising — how many proteins fail each null at `adj_p < 0.05`?

```{r da-summary}
#| eval: true

da_results %>%
  summarise(
    n_total       = dplyr::n(),
    n_time        = sum(adj_p_time        < 0.05, na.rm = TRUE),
    n_fiber_type  = sum(adj_p_fiber_type  < 0.05, na.rm = TRUE),
    n_interaction = sum(adj_p_interaction < 0.05, na.rm = TRUE),
    n_time_up     = sum(adj_p_time < 0.05 & Estimate_time1 > 0, na.rm = TRUE),
    n_time_down   = sum(adj_p_time < 0.05 & Estimate_time1 < 0, na.rm = TRUE)
  )
```

## 2.1 Boxplots of single features

The first sanity check after any DA result is to look at the raw intensities of a few top hits, stratified by the variable of interest.

```{r boxplot-feature}
#| eval: true

# Pick a top hit on the time effect
feature <- da_results %>% arrange(`Pr(>|t|)_time1`) %>% dplyr::slice(1) %>% pull(gene)

plot_df <- tibble(
  fiber_id  = colnames(intensities),
  intensity = intensities[feature, ]
) %>%
  left_join(metadata, by = "fiber_id")

ggplot(plot_df, aes(time, intensity, fill = time)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.3) +
  theme_minimal() +
  labs(title = feature, y = "log2 intensity")
```

> **Exercise 2.1**: Use the example chunk above as a template.
>
> 1.  Pick a protein that is **not significant** for the time test (e.g. one with `adj_p_time > 0.5`) and reproduce the boxplot. Does the lack of a `pre → post` shift match the test statistic?
>
> 2.  Redraw the boxplot of the top time hit (`S100A13`) coloured by `subject_id`, with one line per subject connecting their `pre` and `post` means. Is the shift consistent across all 8 subjects, or driven by a few?

<details>

<summary><strong>Exercise 2.1.1 — solution</strong></summary>

```{r ex-2-1-1}
#| eval: true

nonsig_feature <- da_results %>%
  filter(adj_p_time > 0.5) %>%
  dplyr::slice(1) %>%
  pull(gene)

plot_df <- tibble(
  fiber_id  = colnames(intensities),
  intensity = intensities[nonsig_feature, ]
) %>%
  left_join(metadata, by = "fiber_id")

ggplot(plot_df, aes(time, intensity, fill = time)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.3) +
  theme_minimal() +
  labs(title = nonsig_feature, y = "log2 intensity")
```

Compared to a top hit, the two boxes should overlap heavily and the medians sit on essentially the same line.

</details>

<details>

<summary><strong>Exercise 2.1.2 — solution</strong></summary>

```{r ex-2-1-2}
#| eval: true

feature <- "S100A13"

plot_df <- tibble(
  fiber_id  = colnames(intensities),
  intensity = intensities[feature, ]
) %>%
  left_join(metadata, by = "fiber_id") %>%
  group_by(subject_id, time) %>%
  summarise(intensity = mean(intensity, na.rm = TRUE), .groups = "drop")

ggplot(plot_df, aes(time, intensity, group = subject_id, colour = subject_id)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = feature, y = "log2 intensity (subject mean)")
```

If all 8 lines slope in the same direction, the effect is consistent. If a couple of subjects buck the trend, the population estimate is being pulled by a subset.

</details>

## 2.2 Volcano plots

A volcano plot shows effect size (`Estimate_time1`, the log2FC of `post − pre`) on the x-axis against statistical significance (`-log10 p-value`) on the y-axis. It is the standard one-glance summary of a DA analysis.

```{r volcano}
#| eval: true

# Annotate each protein as significant or not, build a rich hover label,
# then build the volcano in ggplot and pipe it through plotly for hover.
volcano_df <- da_results %>%
  mutate(
    sig   = factor(ifelse(adj_p_time < 0.05, "adj_p < 0.05", "n.s."),
                   levels = c("n.s.", "adj_p < 0.05")),
    label = paste0(
      "gene: ",  gene,
      "<br>log2FC: ", signif(Estimate_time1, 3),
      "<br>adj_p: ", signif(adj_p_time, 3)
    )
  )

p_volcano <- ggplot(
    volcano_df,
    aes(x = Estimate_time1, y = -log10(`Pr(>|t|)_time1`),
        colour = sig, text = label)
  ) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0,            linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = c("n.s." = "grey70", "adj_p < 0.05" = "firebrick")) +
  theme_minimal() +
  labs(x = "log2FC (post − pre)", y = "-log10(p-value)",
       colour = NULL,
       title = "Differential abundance: post vs. pre")

# Static ggplot (useful for reports / printed handouts)
p_volcano

# Interactive plotly version — hover any dot to see gene name and stats
plotly::ggplotly(p_volcano, tooltip = "text")
```

> **Exercise 2.2**:
>
> 1.  Read off the volcano: how many proteins are significant at `adj_p_time < 0.05`, and is the cloud symmetric around `Estimate_time1 = 0`? Use the interactive plotly version — hover over the upper-right and upper-left of the cloud and write down the top few up- and down-regulated genes.
>
> 2.  Redraw the volcano with the same ggplot + plotly template, but using **PC1 loading vs. fiber-type p-value**. PC1 loading lives in `pca_res$rotation[, "PC1"]` — join it onto `da_results` to use it as the x-axis, and put `p_val_fiber_type` on the y-axis. Are there more or fewer hits than in the time volcano?

<details>

<summary><strong>Exercise 2.2.1 — possible answer</strong></summary>

On this dataset, **1,513 / 2,373 proteins (64 %)** have `adj_p_time < 0.05`. The cloud is biased upward — **971 hits go up** (`Estimate_time1 > 0`) vs. 542 down. The perturbation produces a net increase in more proteins than it decreases, suggesting an overall "induction" response rather than a balanced rewiring.

Hovering the upper-right of the plot you should find genes like `S100A13`, `HSPB6`, `CES2`; the upper-left should show down-regulated hits.

</details>

<details>

<summary><strong>Exercise 2.2.2 — solution</strong></summary>

```{r ex-2-2-2}
#| eval: true

# Join PC1 loading onto da_results so we can use it as the x-axis
pc1_load <- tibble(
  gene        = rownames(pca_res$rotation),
  pc1_loading = pca_res$rotation[, "PC1"]
)

volcano_pc1_df <- da_results %>%
  left_join(pc1_load, by = "gene") %>%
  mutate(
    sig   = factor(ifelse(adj_p_fiber_type < 0.05, "adj_p < 0.05", "n.s."),
                   levels = c("n.s.", "adj_p < 0.05")),
    label = paste0(
      "gene: ",         gene,
      "<br>PC1 loading: ", signif(pc1_loading, 3),
      "<br>adj_p: ",       signif(adj_p_fiber_type, 3)
    )
  )

p_volcano_pc1 <- ggplot(
    volcano_pc1_df,
    aes(x = pc1_loading, y = -log10(p_val_fiber_type),
        colour = sig, text = label)
  ) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0,            linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = c("n.s." = "grey70", "adj_p < 0.05" = "firebrick")) +
  theme_minimal() +
  labs(x = "PC1 loading", y = "-log10(adj. p-value)",
       colour = NULL,
       title = "Fiber-type axis (PC1) vs. GAM p-value")

plotly::ggplotly(p_volcano_pc1, tooltip = "text")
```

The fiber-type volcano shows many more hits (**2,223 of 2,373**) than the time volcano — PC1 is the dominant axis in the data, so almost every protein co-varies with it to some degree.Perhaps a volcano plot is not the best way to explore the associations between protein intensities and fiber type, and a heatmap of the associated features across PC1 would be more informative and reveal abundance patterns across the fiber type spectrum.

</details>

## 2.3 Feature plots on UMAP

Overlaying a single protein's intensity onto the UMAP embedding shows *where* in fiber-state space that protein is high or low — often more informative than a boxplot when the contrast is driven by a sub-population.

```{r umap-feature}
#| eval: true

feature <- da_results %>%
    dplyr::filter(gene == "MYH7") |>
    pull(gene)

Seurat::FeaturePlot(seu, features = feature, reduction = "umap",
                    pt.size = 0.4, order = TRUE) +
  ggplot2::scale_colour_viridis_c()
```

> **Exercise 2.3**:
>
> 1.  The example uses a classic **fiber-type** hit. Reproduce the feature plot for the top **time** hit (`S100A13`). Is the protein's expression concentrated in one UMAP region, or does it shift more uniformly across the embedding?
>
> 2.  Repeat for the top **interaction** hit (`ALDH1B1`). Compare to the time-hit map — does the interaction-driven protein localize differently?

<details>

<summary><strong>Exercise 2.3.1 — solution</strong></summary>

```{r ex-2-3-1}
#| eval: true

Seurat::FeaturePlot(seu, features = "S100A13", reduction = "umap",
                    pt.size = 0.4, order = TRUE) +
  ggplot2::scale_colour_viridis_c()
```

`S100A13`'s effect is largely **fiber-type-agnostic** — a top time hit tends to shift across most of the UMAP, not just one cluster.

</details>

<details>

<summary><strong>Exercise 2.3.2 — solution</strong></summary>

```{r ex-2-3-2}
#| eval: true

Seurat::FeaturePlot(seu, features = "ALDH1B1", reduction = "umap",
                    pt.size = 0.4, order = TRUE) +
  ggplot2::scale_colour_viridis_c()
```

`ALDH1B1` is both a top time hit **and** a top interaction hit — so the time-driven shift is concentrated in one fiber-type region rather than spread uniformly. This is the visual signature of a `time × fiber_type` interaction.

</details>

## 2.4 Association of protein intensity with PC1 (fiber type)

`p_val_fiber_type` tests whether a protein's intensity varies along PC1. For a top fiber-type hit, the relationship should be obvious in a scatter of intensity vs. PC1.

```{r intensity-vs-pc1}
#| eval: true

feature <- da_results %>% arrange(p_val_fiber_type) %>% dplyr::slice(1) %>% pull(gene)

plot_df <- tibble(
  fiber_id  = colnames(intensities),
  intensity = intensities[feature, ]
) %>%
  left_join(metadata, by = "fiber_id")

ggplot(plot_df, aes(PC1, intensity)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE) +
  theme_minimal() +
  labs(title = feature, x = "PC1 (fiber-type axis)", y = "log2 intensity")
```

> **Exercise 2.4**:
>
> 1.  Reproduce the scatter for a protein with **very high** `p_val_fiber_type` (e.g. `C4orf54`) and compare it visually to the top hit (`GATD3B`). Does the test result match what you see in the cloud?
>
> 2.  Compute the Pearson correlation between intensity and PC1 for both proteins and confirm.

<details>

<summary><strong>Exercise 2.4.1 — solution</strong></summary>

```{r ex-2-4-1}
#| eval: true

plot_one <- function(gene) {
  plot_df <- tibble(
    fiber_id  = colnames(intensities),
    intensity = intensities[gene, ]
  ) %>%
    left_join(metadata, by = "fiber_id")

  ggplot(plot_df, aes(PC1, intensity)) +
    geom_point(size = 0.4, alpha = 0.4) +
    geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE) +
    theme_minimal() +
    labs(title = gene, x = "PC1", y = "log2 intensity")
}

plot_one("GATD3B")   # top fiber-type hit
plot_one("C4orf54")  # not significant for fiber type
```

`GATD3B` shows a clean monotonic gradient with PC1; `C4orf54`'s cloud is flat.

</details>

<details>

<summary><strong>Exercise 2.4.2 — solution</strong></summary>

```{r ex-2-4-2}
#| eval: true

cor_with_pc1 <- function(gene) {
  vals <- intensities[gene, ]
  ok   <- !is.na(vals)
  cor(vals[ok], metadata$PC1[ok])
}

c(
  GATD3B  = cor_with_pc1("GATD3B"),
  C4orf54 = cor_with_pc1("C4orf54")
)
```

`GATD3B` should give |cor| roughly between 0.6 and 0.9; `C4orf54` should be close to 0.

</details>

## 2.5 Interaction between PC1 and time

`p_val_interaction` flags proteins whose PC1 dependence **differs between pre and post** — i.e. the time effect is fiber-type-specific. The cleanest visualization is intensity vs. PC1, split by time:

```{r interaction-plot}
#| eval: true

feature <- da_results %>% arrange(p_val_interaction) %>% dplyr::slice(1) %>% pull(gene)

plot_df <- tibble(
  fiber_id  = colnames(intensities),
  intensity = intensities[feature, ]
) %>%
  left_join(metadata, by = "fiber_id")

ggplot(plot_df, aes(PC1, intensity, colour = time, fill = time)) +
  geom_point(size = 0.4, alpha = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE, alpha = 0.2) +
  theme_minimal() +
  labs(title = feature, x = "PC1 (fiber-type axis)", y = "log2 intensity",
       subtitle = "Two curves with different slopes => significant interaction")
```

> **Exercise 2.5**:
>
> 1.  Make the same `intensity ~ PC1` plot, split by `time`, for a protein that has a **strong time effect but no interaction** — e.g. `S100A13` (top time hit, weak interaction). Are the two smoothers roughly parallel?
>
> 2.  Repeat for `ALDH1B1`, which is a top hit for **both** time and interaction. Do the two smoothers cross or just diverge? What does that pattern imply biologically?

<details>

<summary><strong>Exercise 2.5.1 — solution</strong></summary>

```{r ex-2-5-1}
#| eval: true

plot_interaction <- function(gene) {
  plot_df <- tibble(
    fiber_id  = colnames(intensities),
    intensity = intensities[gene, ]
  ) %>%
    left_join(metadata, by = "fiber_id")

  ggplot(plot_df, aes(PC1, intensity, colour = time, fill = time)) +
    geom_point(size = 0.4, alpha = 0.3) +
    geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE, alpha = 0.2) +
    theme_minimal() +
    labs(title = gene, x = "PC1", y = "log2 intensity")
}

plot_interaction("S100A13")
```

For a pure time effect, the two smoothers should sit at different `y`-levels but track the same shape — vertically shifted but **parallel**. That visual is the signature of "time effect, no interaction".

</details>

<details>

<summary><strong>Exercise 2.5.2 — solution</strong></summary>

```{r ex-2-5-2}
#| eval: true

plot_interaction("ALDH1B1")
```

`ALDH1B1` is a top hit for both `Pr(>|t|)_time1` and `p_val_interaction`. The two smoothers should be clearly **non-parallel** — the gap between pre and post widens (or narrows) along PC1.

Biologically: a significant interaction means the `post − pre` change is not uniform across fibers — one fiber-type tail responds more than the other. If the curves *cross*, the direction of the time effect flips between fiber types; if they only *diverge*, one fiber type responds and the other barely does. Either pattern is what triggers a small `p_val_interaction` in the GAM.

</details>

------------------------------------------------------------------------

# Stage 3: Heatmaps, Co-Regulated Modules & Pathway Enrichment

A volcano plot tells you *which* proteins change, but not whether they change *together*. A clustered heatmap of differentially abundant proteins reveals **modules** — groups of proteins that follow the same expression pattern across fibers — which often correspond to a shared biological process.

## 3.1 Heatmap of differentially abundant proteins

```{r heatmap}
#| eval: true

sig_proteins <- da_results %>%
  filter(adj_p_time < 0.05) %>%
  pull(gene)

mat <- intensities[sig_proteins, ]
mat <- t(scale(t(mat)))   # z-score per protein (row-wise)

# Mean-impute remaining NAs for clustering
row_means <- rowMeans(mat, na.rm = TRUE)
na_idx <- which(is.na(mat), arr.ind = TRUE)
mat[na_idx] <- row_means[na_idx[, "row"]]

# Order columns by PC1 so the fiber-type axis reads left-to-right
col_order <- order(metadata$PC1)
mat <- mat[, col_order]

# pheatmap reads annotations from a data frame whose rownames match colnames(mat)
ann_col <- metadata %>%
  dplyr::select(fiber_id, time, PC1) %>%
  column_to_rownames("fiber_id")
ann_col <- ann_col[colnames(mat), , drop = FALSE]

ph <- pheatmap::pheatmap(
  mat,
  annotation_col   = ann_col,
  show_rownames    = FALSE,
  show_colnames    = FALSE,
  cluster_rows     = TRUE,
  cluster_cols     = FALSE,   # columns already ordered by PC1
  cutree_rows      = 6,       # <- number of protein modules, tune
  scale            = "none",  # already z-scored above
  fontsize         = 8
)
```

> **Exercise 3.1**: Increase `row_km` from 4 to 10. At what value do additional modules become redundant (i.e. two splits that show essentially the same pattern)? This is a practical heuristic for the number of meaningful modules.

<details>

<summary><strong>Exercise 3.1 — possible answer</strong></summary>

TODO.

</details>

## 3.2 Extracting module membership

```{r modules}
#| eval: true

# pheatmap returns the row hclust as `tree_row`; cut it at the same k used above
module_id <- cutree(ph$tree_row, k = 6)
modules   <- tibble(protein = names(module_id), module = unname(module_id))

modules %>% count(module)
```

## 3.3 Pathway enrichment per module

For each module, test whether its protein set is enriched for any annotated pathway (Reactome, KEGG, GO BP). Tools: `clusterProfiler`, `fgsea`, `enrichR`.

```{r enrichment}
#| eval: true

# Per-module over-representation analysis
# enrich_res <- modules %>%
#   group_split(module) %>%
#   map(~ clusterProfiler::enrichGO(
#         gene         = .x$protein,
#         universe     = rownames(intensities),
#         keyType      = "SYMBOL",
#         OrgDb        = org.Hs.eg.db::org.Hs.eg.db,
#         ont          = "BP",
#         pAdjustMethod = "BH"
#       ))
```

> **Exercise 3.2**: For the module most strongly associated with the time effect, what is the top enriched pathway? Does it match the known biology of the perturbation?

<details>

<summary><strong>Exercise 3.2 — possible answer</strong></summary>

TODO.

</details>

------------------------------------------------------------------------

# Extended Exercises

### Exercise A — PCA input choice

Re-run the PCA after (i) keeping all proteins, (ii) keeping only the top-2000 most variable proteins, and (iii) keeping only the differentially abundant proteins. How does the variance explained by PC1 change? Which input gives the cleanest separation of `time`?

### Exercise B — UMAP stability

UMAP is stochastic. Run UMAP three times with different `set.seed()` values and overlay the results. Are the global structures (clusters) consistent? Are the local positions of individual fibers consistent?

### Exercise C — Volcano cutoffs

Re-plot the volcano with `FCcutoff = 0.5`, `1`, and `2`. How does the count of "significant" proteins change? When is the FC cutoff justified by the biology vs. arbitrary?

### Exercise D — Module robustness

Bootstrap the heatmap: sample 80 % of the fibers without replacement, redo the clustering, and check whether the same modules emerge. This gives an empirical measure of module stability.

### Exercise E — From modules to mechanism

Pick one enriched pathway from Exercise 3.2 and find a curated regulator of that pathway (transcription factor, kinase). Is the regulator itself differentially abundant? If not, does it lie in any of the modules?
